{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-26 thesis/landslide.tex
nobe

The execution tree
%(or state space, or preemption point log, however calling it suits you)
is stored as a chain of preemption point nodes named {\tt struct hax} defined in {\tt tree.h}.
Although the state space of possible interleavings is exponentially-sized,
Landslide does not actually need to store any nodes for execution sequences outside the current variation
(see \cref{sec:landslide-estimate} and \cref{sec:landslide-dpor} for why),
so the total memory consumption is only $O(n)$ in the number of preemption points in a single program run
(for the test cases used in this thesis, typically 20-1000).
Each {\tt hax} stores the following information:

@@

In Bochs, {\tt fork()} automatically copies everything, so the reverse holds:
all data of the latter case must whenever updated be propagaged to all {\tt fork()}ed children processes explicitly.
I worried while implementing this that I might miss a case, or that future updates to the code could easily forget this step,
resulting undoubtedly in state corruption bugs which to diagnose would be a thesis in their own right,
so I enlisted help from my compiler via the oft-ridiculed {\tt const}.
Every preemption point node in the execution tree ({\tt tree.h}),
each of whose state is kept generally read-only,
and all modifications must go through {\tt modify\_hax()} ({\tt timetravel.h}) using a modification callback,
which internally casts away the {\tt const}, performs the requested modification,
and also messages all relevant child processes to perform the same ({\tt timetravel\_set()} in {\tt timetravel.c}).
The {\tt const} is absolutely, inviolably, not to be casted away, at the sacred cost of what little type safety C offers.%
\footnote{Of course this would be followed by a footnote describing the one place where I cast it away anyway,
{\tt mem\_check\_shared\_access()} in {\tt memory.c};
why it's ok is documented in an {\tt XXX} comment in the code.}
Thence the typechecker enforces that all exploration-related state is properly propagated while scheduler state is automatically reverted.

@@

Whenever {\tt arbiter\_interested()} returns true,
Landslide creates a new {\tt struct hax} in the execution tree (\cref{sec:landslide-save}),
creates a checkpoint (\cref{sec:landslide-timetravel}),
and queries {\tt arbiter\_choose()} to decide which thread to run next (\cref{sec:landslide-arbiter}).

@@

{\bf Definition.}
Given a transition sequence (execution, interleaving, preemption trace) $S = t_1 \dots t_n$,
the DPOR algorithm identifies any number of alternate interleavings that must be tested.
Each such interleaving I will denote in this section as $I_{ij} = (t_1 \dots t_{i-1}, T_j)$,
where $t_1 \dots t_{i-1}$ is the common execution prefix shared between $S$ and the new interleaving,
and $T_j$ is the thread ID to be scheduled after $t_{i-1}$, $T_j \ne T(t_i)$.
%to run {\em instead} of $T(t_i)$.
\footnote{I describe $T_j$ as a thread ID here, rather than as a thread transition,
because the nature of the transition (its memory accesses, the subsequent state, et cetera)
is unknown until actually executed.}
Landslide's implementation represents $S$ as a list of {\tt struct hax}es, defined in {\tt tree.h},
each one representing a preemption point, or intermediate state between two transitions.

@@

{\bf Tracking already-visited interleavings.}
Let $U(I_{ij})$ denote the sub-state-space (or sub-tree) beginning at the next preemption point reached
after executing $T_j$ after $t_1 \dots t_{i-1}$;
in other words, the set of all sequences $S' = t_1 \dots t_{i-1}, u_i, \dots u_{n'}$ with $T(u_i) = T_j$.
Landslide orders its search depth-first,
so for any such $U$ outside the current interleaving,
either all or none of its $S'$s will have been tested already.
Therefore, to avoid repeating interleavings,
Landslide need only store at each {\tt struct hax} a list of threads
such that their corresponding subtree $U$ is fully explored,
and can omit any non-constant-size information about the contents of that $U$
({\tt struct hax\_child}).
Hence the memory cost of Landslide's DPOR implementation is $O(nk)$,
$k$ being the maximum runnable threads at any preemption point
(which in turn is always single digits for model checking tests).

{\bf Choosing which new interleaving to test next.}
Among all interleavings chosen by DPOR not already marked as explored,
Landslide chooses the one with the longest execution prefix matching the current $S$,
to maintain the depth-first search invariant.
(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
All other new interleavings are marked to explore later, %in their corresponding {\tt struct hax},
and automatically included in the result of any future iterations of DPOR until they are tested.
Because the one with the longest execution prefix was chosen to test next,
all others must share their execution prefixes with it,
preserving the $O(nk)$ memory bound described above.
%so storing markers for the others during the next execution is still $O(1)$ memory.

>>>>

The execution tree
%(or state space, or preemption point log, however calling it suits you)
is stored as a chain of preemption point nodes named {\tt struct nobe} defined in {\tt tree.h}.
Although the state space of possible interleavings is exponentially-sized,
Landslide does not actually need to store any nodes for execution sequences outside the current variation
(see \cref{sec:landslide-estimate} and \cref{sec:landslide-dpor} for why),
so the total memory consumption is only $O(n)$ in the number of preemption points in a single program run
(for the test cases used in this thesis, typically 20-1000).
Each {\tt nobe} stores the following information:

@@

In Bochs, {\tt fork()} automatically copies everything, so the reverse holds:
all data of the latter case must whenever updated be propagaged to all {\tt fork()}ed children processes explicitly.
I worried while implementing this that I might miss a case, or that future updates to the code could easily forget this step,
resulting undoubtedly in state corruption bugs which to diagnose would be a thesis in their own right,
so I enlisted help from my compiler via the oft-ridiculed {\tt const}.
Every preemption point node in the execution tree ({\tt tree.h}),
each of whose state is kept generally read-only,
and all modifications must go through {\tt modify\_nobe()} ({\tt timetravel.h}) using a modification callback,
which internally casts away the {\tt const}, performs the requested modification,
and also messages all relevant child processes to perform the same ({\tt timetravel\_set()} in {\tt timetravel.c}).
The {\tt const} is absolutely, inviolably, not to be casted away, at the sacred cost of what little type safety C offers.%
\footnote{Of course this would be followed by a footnote describing the one place where I cast it away anyway,
{\tt mem\_check\_shared\_access()} in {\tt memory.c};
why it's ok is documented in an {\tt XXX} comment in the code.}
Thence the typechecker enforces that all exploration-related state is properly propagated while scheduler state is automatically reverted.

@@

Whenever {\tt arbiter\_interested()} returns true,
Landslide creates a new {\tt struct nobe} in the execution tree (\cref{sec:landslide-save}),
creates a checkpoint (\cref{sec:landslide-timetravel}),
and queries {\tt arbiter\_choose()} to decide which thread to run next (\cref{sec:landslide-arbiter}).

@@

{\bf Definition.}
Given a transition sequence (execution, interleaving, preemption trace) $S = t_1 \dots t_n$,
the DPOR algorithm identifies any number of alternate interleavings that must be tested.
Each such interleaving I will denote in this section as $I_{ij} = (t_1 \dots t_{i-1}, T_j)$,
where $t_1 \dots t_{i-1}$ is the common execution prefix shared between $S$ and the new interleaving,
and $T_j$ is the thread ID to be scheduled after $t_{i-1}$, $T_j \ne T(t_i)$.
%to run {\em instead} of $T(t_i)$.
\footnote{I describe $T_j$ as a thread ID here, rather than as a thread transition,
because the nature of the transition (its memory accesses, the subsequent state, et cetera)
is unknown until actually executed.}
Landslide's implementation represents $S$ as a list of {\tt struct nobe}s, defined in {\tt tree.h},
each one representing a preemption point, or intermediate state between two transitions.

@@

{\bf Tracking already-visited interleavings.}
Let $U(I_{ij})$ denote the sub-state-space (or sub-tree) beginning at the next preemption point reached
after executing $T_j$ after $t_1 \dots t_{i-1}$;
in other words, the set of all sequences $S' = t_1 \dots t_{i-1}, u_i, \dots u_{n'}$ with $T(u_i) = T_j$.
Landslide orders its search depth-first,
so for any such $U$ outside the current interleaving,
either all or none of its $S'$s will have been tested already.
Therefore, to avoid repeating interleavings,
Landslide need only store at each {\tt struct nobe} a list of threads
such that their corresponding subtree $U$ is fully explored,
and can omit any non-constant-size information about the contents of that $U$
({\tt struct nobe\_child}).
Hence the memory cost of Landslide's DPOR implementation is $O(nk)$,
$k$ being the maximum runnable threads at any preemption point
(which in turn is always single digits for model checking tests).

{\bf Choosing which new interleaving to test next.}
Among all interleavings chosen by DPOR not already marked as explored,
Landslide chooses the one with the longest execution prefix matching the current $S$,
to maintain the depth-first search invariant.
(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
All other new interleavings are marked to explore later, %in their corresponding {\tt struct nobe},
and automatically included in the result of any future iterations of DPOR until they are tested.
Because the one with the longest execution prefix was chosen to test next,
all others must share their execution prefixes with it,
preserving the $O(nk)$ memory bound described above.
%so storing markers for the others during the next execution is still $O(1)$ memory.

### 2018-12-26 thesis/tm.tex
nobe

Each preemption point ({\tt struct hax} as defined in {\tt tree.h})
includes three new fields.
The boolean {\tt h->xbegin} is set if the preemption point occurred at an {\tt \_xbegin()} call.
Then if set (as the tag of an option type),
the twin lists of integers
{\tt h->xabort\_codes\_ever}
and
{\tt h->xabort\_codes\_todo}
store possible error codes this {\tt \_xbegin()} call should be tested to possibly return.
The former list stores all such error codes, whether already tested or yet to be tested,
while the latter serves as a workqueue that indicates only those not already tested yet
(serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
The state space estimators (\cref{sec:landslide-estimate})
check the length of these lists, in addition to DPOR's tagged threads,
when computing the number of marked children.

@@

For all the reduction retry sets demonstrated in \Cref{tab:tm-verifs},
some inefficiencies remain in its strategy.
For example, it is not clear how to prune soundly
when three or more threads must be reorderd around one transactional preemption point,
or when a second pair of partially-independent transactions interleaves while an existing retry set is already active.
Accordingly, I implemented the optimization as conservatively as possible in these cases,
``saturating'' the retry sets to fall back to no pruning
({\tt update\_hax\_abort\_set()} and
{\tt update\_hax\_abandon\_abort\_set()}, respectively).

>>>>

Each preemption point ({\tt struct nobe} as defined in {\tt tree.h})
includes three new fields.
The boolean {\tt h->xbegin} is set if the preemption point occurred at an {\tt \_xbegin()} call.
Then if set (as the tag of an option type),
the twin lists of integers
{\tt h->xabort\_codes\_ever}
and
{\tt h->xabort\_codes\_todo}
store possible error codes this {\tt \_xbegin()} call should be tested to possibly return.
The former list stores all such error codes, whether already tested or yet to be tested,
while the latter serves as a workqueue that indicates only those not already tested yet
(serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
The state space estimators (\cref{sec:landslide-estimate})
check the length of these lists, in addition to DPOR's tagged threads,
when computing the number of marked children.

@@

For all the reduction retry sets demonstrated in \Cref{tab:tm-verifs},
some inefficiencies remain in its strategy.
For example, it is not clear how to prune soundly
when three or more threads must be reorderd around one transactional preemption point,
or when a second pair of partially-independent transactions interleaves while an existing retry set is already active.
Accordingly, I implemented the optimization as conservatively as possible in these cases,
``saturating'' the retry sets to fall back to no pruning
({\tt update\_nobe\_abort\_set()} and
{\tt update\_nobe\_abandon\_abort\_set()}, respectively).
